---
output: github_document
title: "SMOPPIX: Single-MOlecule sPatial omics data analysed using the Probabilistic IndeX"
author: Stijn Hawinkel
---

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("inst/Smoppix.jpg"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width = "160px",
               heigth = "160px")
```

This repo provides code for analyzing spatial transcriptomics and other omics data on the single-molecule level using probabilistic indices. A simple use-case is shown below, more extensive documentation can be found in the vignette

The package can be installed from GitHub as follows:

```{r installAndLoadGitHub, eval = FALSE}
library(devtools)
install_github("sthawinke/smoppix")
```

```{r install, eval = FALSE}
library(Biocmanager)
install("smoppix")
```

Load the package

```{r loadSmoppix}
library(smoppix)
```

Load a dataset, convert it to a _spatstat_ hyperframe, and make an exploratory plot:

```{r loadYang}
data(Yang)
hypYang <- buildHyperFrame(Yang,
    coordVars = c("x", "y"),
    imageVars = c("day", "root", "section")
)
plotExplore(hypYang)
```

Estimate the univariate nearest neighbour probabilistic index:

```{r nn}
nnObj <- estPis(hypYang, pis = "nn", null = "background", verbose = FALSE)
```

Add a variance weighting function and plot it

```{r wf}
nnObj <- addWeightFunction(nnObj, designVars = c("day", "root"))
plotWf(nnObj, pi = "nn")
```

The inference step: fit linear mixed models, and show the most significant results:

```{r inference}
allModsNN <- fitLMMs(nnObj, fixedVars = "day", randomVars = "root")
head(getResults(allModsNN, "nn", "Intercept"))
```

Let's make it visual and plot the most significantly aggregated transcripts:

```{r plotTopRes}
plotTopResults(hypYang, allModsNN, pi = "nn")
```

Finally write the results to a spreadsheet:

```{r writeToSpreadsheet, eval = FALSE}
writeToXlsx(allModsNN, file = "myfile.xlsx")
```
